Static contract checking for Haskell

### 2011-10-26 paper.tex
Paper: formatting.
Paper: expanded prelude and contract translation sections, including figures.
Put the new contract translation presentation in a figure, and wrote a few paragraphs
explaining the motivation behind the min()s.

@@

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
  text.  The lowercase $true$ and $false$ is an artifact of the implementation, i.e. Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.  Moreover, the lower case here is inconsistent with the rest of the presentation, e.g. $Cons$ and $Nil$ above.}
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
\[
notnull(Nil) = false
\]? The answer is that this is an expansion of the $etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.  I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.  In this special case we could remove the implication, but not in general.  The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}

\begin{equation}
  \label{defnotnull}
\forall xs,y,ys.~ x=Cons(y,ys) \implies notnull(xs) = true
\end{equation}

We also need to specify the translation of calls to $notnull$ with
$\bad$ values (to encode the fact that $notnull(\bad) = \bad$.
\begin{equation*}
\forall xs.~xs=\bad \to notnull(xs) = \bad
\end{equation*}

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $phi$, $\forall x,y.x=y \to phi \iff phi[y/x]$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}

@@

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
    ``type'' is used intuitively here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a\text{\note[nc]{This was $K_i(\vec{x})$.  Typo, right?}}}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

\subsection{$\ctrans{}$ -- Contracts}
We give in figure~\ref{ctrans} the translation of contract satisfaction.
\annote[nc]{$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.}{
Again, we should use $True$ (or \texttt{True}) in the translation; the
lowercase is an artifact of the concrete TPTP syntax.
}
\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{

@@

-- what we actually discussed was adding a top level min(e) assumption, but
-- removing the min(e) goal has a similar effect.
[[e in cf]]       :=           cf(e)
not[[e notin cf]] := min(e) -> cf(e)
\end{verbatim}
\note[nc]{And what do we see?
If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
then the ``new'' translation looks almost the same,
except it has more $\min{}$'s in the predicate rules.  So, have we learned
anything? Probably:
\protect \begin{enumerate}
  \item We're more confident about our suggested changes to the old translation,
        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
  \item The new translation is more intuitive / we might understand how Koen came up
        with the old translation.
  \item We can answer the question about what we give as input to Equinox: for the
        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
        That is consistent with giving Equinox $\in$ axioms.  
        What to give as the goal is maybe
        less obvious, since Equinox is a refutation prover, but consider: ignoring the
        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
        But
        that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.  
        And going back
        to the old translation, the analogy is 
        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
\end{enumerate}

@@

\subsection{Prelude}
There are some formulae that should always be included in our FO theory.

We need to state that $\bad$ is not crash-free with the formula:
$\lnot \cf{\bad}$.

\add[nc]{We also state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{

This is already present in the implementation, and assumed in the
consistency proof.}

Plus we need to give formulae for the boolean datatype and for
 unreachability. Strictly speaking, we can omit them and just add the
following lines to source files:
\begin{verbatim}
data UNR = UNR
data Bool = True | False
\end{verbatim}
But given that those datatypes are used by our translation, we can
just directly include their translation every time we translate a
module.

>>>>

\newcommand{\designChoice}{(DESIGN CHOICE/OPTIONAL)}

@@

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
  text.  The lowercase $true$ and $false$ is an artifact of the implementation, i.e. Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.  Moreover, the lower case here is inconsistent with the rest of the presentation, e.g. $Cons$ and $Nil$ above.}
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
\[
notnull(Nil) = false
\]? The answer is that this is an expansion of the $\etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.  I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.  In this special case we could remove the implication, but not in general.  The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}

\begin{equation}
  \label{defnotnull}
\forall xs,y,ys.~ xs=Cons(y,ys) \implies notnull(xs) = true
\end{equation}

We also need to specify the translation of calls to $notnull$ with
$\bad$ values, to encode the fact that $notnull(\bad) = \bad$.
\begin{equation*}
\forall xs.~xs=\bad \to notnull(xs) = \bad
\end{equation*}

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\note[nc]{I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.}
\note[nc]{See Section~\ref{sec:contracts} for a discussion of the $\min{}$ predicate.}

@@

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1, \dots, K_n} = \Phi_1 \cup \Phi_2 \cup \Phi_3 \cup \Phi_4$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \Phi_1 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to \bigwedge_{1 \leq j \leq k} x_j = sel_{j,K_i}(a) \\
      \Phi_2 &=& \bigcup_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = K_i(\vec{x}) \land a = K_j(\vec{x})) \\
      \Phi_3 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \Phi_4 &=& \bigcup_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

\subsection{$\ctrans{}$ -- Contracts}\label{sec:contracts}
\note[nc]{%
Figure~\ref{fig:newCTrans} contains the new min contract translation, and
Figure~\ref{fig:naiveCTrans} contains the naive contract translation
with no $\min$'s.  The min translation reduces to the naive translation when
all $\min$'s are assumed true, i.e. when an axiom $\forall x.\min{x}$ is added.

The min translation has two forms, $\ctrans{}^-$ for 
assumptions/axioms and $\ctrans{}^+$
for conclusions/theorems.  The choice of minus and 
plus comes from the negative and positive
positions in an implication: $- \to +$.  The goal of the $\min$'s is to limit the
applicability of the axioms ($\ctrans{}^-$ and $\dtrans{}$ and $\ktrans{}$), 
in order to prune 
the search space.  So, $\min$'s
appear as obligations (the premises in implications) 
in contract and definition axioms. The mnemonic is ``(a)m in(terested)'',
and the translation is designed so that $\min{e}$ is introduced whenever
$e$ is evaluated or inspected.  The axiom translations contain $\min{}$
obligations that prevent using the axioms on ``uninteresting'' terms.

For example, consider $\ctrans{e \in \cfc}^+$ and $\ctrans{e \in \cfc}^-$
in Figure~\ref{fig:newCTrans}.  In the plus-translation, $\min{e}$ is a
premise that can be used to prove $\cf{e}$, whereas in the minus-translation, either
there is no $\min{e}$, or $\min{e}$ is an obligation that must be
satisfied before $\cf{e}$ may be concluded.  In the plus-translation, we
get interest ($\min{e}$) in $e$ because we are trying to prove something about $e$,
whereas in the minus-translation, we might have to prove we're interested in $e$ before
we can conclude that $e$ is crash free (this latter $\min{e}$ obligation is
a design choice because it's not clear why we'd want to constrain the use
of $\cf{e}$: we'd need to prove $\min{e}$ anyway before instantiating 
another axiom at $e$).

As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
in Figure~\ref{dtrans}.  Definition translations are axioms, 
so the the initial $\min{f(\vec{x})}$
is an obligation: we aren't allowed to 
``unfold the definition'' of the function unless we prove
that we're interested in a particular application.  
Once that obligation is satisfied, interest ($\min{e}$) in the
scrutinee of the case match may be concluded, because the function
must inspect the scrutinee to compute the result.

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{True,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{True,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}
\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}

[[e in cf]]+ :=  min(e) ->  cf(e)
[[e in cf]]- := [min(e) ->] cf(e)

-- 'min's optional if a
-- 'forall f,x. min(f x) -> min(f)' axiom
-- is added to prelude.
[[e in x:c1 -> c2]]+
  := forall x. [min(e x) ->] [[x in c1]]- -> [[e x in c2]]+   
[[e in x:c1 -> c2]]-
  := forall x. [min(e x) ->] [[x in c1]]+ -> [[e x in c2]]-
\end{verbatim}
\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
of optional parts, are enclosed in single square brackets.}
\label{fig:newCTrans}
\end{figure*}

\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
[[e in cf]]         := cf(e)
[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
\end{verbatim}
\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
\label{fig:naiveCTrans}
\end{figure*}

We give in figure~\ref{ctrans} the translation of contract satisfaction.
\annote[nc]{$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.}{
Again, we should use $True$ (or \texttt{True}) in the translation; the
lowercase is an artifact of the concrete TPTP syntax.
}
\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{

@@

-- what we actually discussed was 
-- adding a top level min(e) assumption, but
-- removing the min(e) goal has a similar effect.
[[e in cf]]       :=           cf(e)
not[[e notin cf]] := min(e) -> cf(e)
\end{verbatim}
\note[nc]{And what do we see?
If we identify $\in^+$ with $\lnot\notin$ and $\in^-$ with $\in$, 
then the ``new'' translation looks almost the same,
except it has more $\min{}$'s in the predicate rules.  So, have we learned
anything? Probably:
\protect \begin{enumerate}
  \item We're more confident about our suggested changes to the old translation,
        including adding a symmetric $\min{}$ assumption to the $\notin$ arrow case. 
  \item The new translation is more intuitive / we might understand how Koen came up
        with the old translation.
  \item We can answer the question about what we give as input to Equinox: for the
        new translation, it's obvious that the theory $T$ should consist of $\in^-$ axioms.
        That is consistent with giving Equinox $\in$ axioms.  
        What to give as the goal is maybe
        less obvious, since Equinox is a refutation prover, but consider: ignoring the
        strategy (refutation), we want to prove $\Gamma^{\in^-} \vdash \ctrans{e \in c}^+$.  
        But
        that's equivalent to $\Gamma^{\in^-},\lnot\ctrans{e \in c}^+ \vdash \bot$.  
        And going back
        to the old translation, the analogy is 
        $\Gamma^{\in}, \ctrans{e \notin c} \vdash \bot$.
\end{enumerate}

@@

\subsection{Prelude}
Figure~\ref{fig:prelude} summarizes the common axioms we add to all
translations.  We 
\begin{itemize}
  \item state that $\bad$ is not crash-free with the formula:
$\lnot \cf{\bad}$.

  \item \add[nc]{state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
This is already present in the implementation, and assumed in the
consistency proof.}

  \item give formulae for the boolean datatype, equivalent to
translating
\begin{verbatim}
data Bool = True 0 | False 0
\end{verbatim}

  \item \designChoice state that interest in an application implies
  interest in the function: $\forall f,x. \min{f ~x} \to \min{f}$.  This axiom can
  be used to eliminate some $\min{}$'s in the translations for arrow contracts.

  \item \designChoice state that applying $\bad$ and $\unr$ as functions
  results in $\bad$ and $\unr$, respecitively: \[
  \forall x. \min{\bad ~x} \to \bad ~x = \bad \land \min{\unr ~x} \to \unr ~x = \unr.
  \]

  \item \designChoice state that applying a $\cfc$ expression can be used like a
  $\cfc \to \cfc$ function: $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$.
  Simon suggested this to allow one to write, e.g., $f ::: \cfc$ instead of 
  $f ::: \cfc \to \cfc$.
\end{itemize}

\begin{figure}\label{fig:prelude}
  \begin{itemize}
    \item $\lnot \cf{\bad}$
    \item $\cf{\unr}$
    \item $\dtrans{data~Bool = True~0 | False~0}$
    \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$
    \item \designChoice\\ $\forall x. \min{\bad ~x} \to \bad ~x = \bad$ 
    \item \designChoice\\ $\forall x. \min{\unr ~x} \to \unr ~x = \unr$
    \item \designChoice\\ $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$
    \item 
  \end{itemize}
  \caption{Prelude -- Axioms common to all translations.}
\end{figure}
